Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-04-08 article.tex
Hernan's changes

For COPASI and Tellurium,
we noted that its runtime depended on the number of intermediate evaluation points (Figure~\ref{fig-runtime}, left).
For SimBio,
its NumPy backend can be orders of magnitude slower than both COPASI and RoadRunner  (Figure~\ref{fig-runtime}, right).
Nevertheless, switching to the numba backend,
which \ac{JIT} compiles the \ac{RHS} equations,
puts it on par with them.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids calling into the Python interpreter
between each of the integration steps.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.
For the models considered,
amortizing the compilation time required from 2 upto 200 runs.

@@

The inclusion of \ac{SBML} support facilitates the effortless reuse of models created by the systems biology community,
along with the vast collection of public models hosted in the BioModels repository.
The modular architecture of these packages facilitates their reuse, enhancement, and extension by the wider Python community.
For instance, an individual from outside the systems biology field could contribute a stochastic integrator to poincaré,
which would then be available in SimBio.
This clear separation of concerns also makes the packages more comprehensible,
lowering the barrier for contributing improvements or new features.
Such an architecture ensures their maintainability and ongoing development well into the future.

@@

Source code repositories for these packages can be found at
\url{https://github.com/maurosilber/poincare} and
\url{https://github.com/hgrecco/simbio}.

>>>>

For COPASI and Tellurium,
we noted that its runtime depended on the number of intermediate evaluation points returned to the user (Figure~\ref{fig-runtime}, left).
We speculate that this is due to memory allocation and data transfer in the Python bindings, 
as the number of total function evaluations (i.e. including those performed by the integrator steper)
is around 3000 and therefore much larger than most part of the x-axis.
For SimBio,
its NumPy backend can be orders of magnitude slower than both COPASI and RoadRunner  (Figure~\ref{fig-runtime}, right).
Nevertheless, switching to the numba backend,
which \ac{JIT} compiles the \ac{RHS} equations,
puts it on par with them.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids calling into the Python interpreter
between each of the integration steps.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.
For the models considered,
amortizing the compilation time required from 2 upto 200 runs.

@@

The inclusion of \ac{SBML} support facilitates the effortless reuse of models created by the systems biology community,
along with the vast collection of public models hosted in the BioModels repository.
The modular architecture of these packages facilitates their reuse, enhancement, and extension by the wider Python community.
Therefore, it should be also easy to integrate with existing infrastructure such a BioSimulators.org  \cite{shaikhBioSimulatorsCentralRegistry2022},
or combine with other packages like SimService \cite{sego_simservice_2024} to build more complex simulations.
For instance, an individual from outside the systems biology field could contribute a stochastic integrator to poincaré,
which would then be available in SimBio.
This clear separation of concerns also makes the packages more comprehensible,
lowering the barrier for contributing improvements or new features.
Such an architecture ensures their maintainability and ongoing development well into the future.

@@

Source code repositories, example and documentation for these packages can be found at
\url{https://github.com/maurosilber/poincare} and
\url{https://github.com/hgrecco/simbio}.
